---
title: "Exercise02"
author: "Immanuel Klein"
format: html
editor: visual
---

```{r}
library("tidyverse")
library("ggplot2")
```

# 2 Examine the distribution

Reading the data:

```{r}
#Reading the whole data set
winequality.white <- read.csv("wine+quality/winequality-white.csv", sep = ";")

#Using only volatile.acidity, residual.sugar, and pH and adding binary variable good which is 1 if quality > 5 and 0 otherwise.
working.df <- winequality.white %>%
  mutate(good = ifelse(quality > 5, 1, 0)) %>%
  select(volatile.acidity, residual.sugar, pH, good)

head(working.df)
```

## Exercise (a)

```{r}
pH.mean <- working.df %>% pull(pH) %>% mean()
pH.sd <- working.df %>% pull(pH) %>% sd()

pH.good.mean <- working.df %>% filter(good == 1) %>% pull(pH) %>% mean()
pH.good.sd <- working.df %>% filter(good == 1) %>% pull(pH) %>% sd()

pH.bad.mean <- working.df %>% filter(good == 0) %>% pull(pH) %>% mean()
pH.bad.sd <- working.df %>% filter(good == 0) %>% pull(pH) %>% sd()

# Plot histogram of pH for all wines
ggplot(working.df, aes(x = pH)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30, alpha = 0.5) + 
  labs(title = "Histogram of pH", 
       x = "pH", 
       y = "Frequency") + 
  stat_function(fun = dnorm, 
                args = c(mean = pH.mean, sd = pH.sd), 
                color = "red")

ggplot(working.df) +
  aes(x = pH, fill = factor(good)) +
  geom_histogram(aes(y = after_stat(density)), 
                 position = "identity", 
                 alpha = 0.5, 
                 bins = 30) + 
  labs(title = "Histogram of pH", 
       x = "pH", 
       y = "Frequency", 
       fill = "Good Wine") +
  stat_function(fun = dnorm, 
                args = c(mean = pH.good.mean, sd = pH.good.sd), 
                color = "red") +
  stat_function(fun = dnorm, 
                args = c(mean = pH.bad.mean, sd = pH.bad.sd), 
                color = "blue")
```

The distributions of good and bad wines deviate from their normal distributions in the following ways:

-   **Good Wines:** The histogram is quite similar to the normal distribution, suggesting that the pH values of good wines are approximately normally distributed.

-   **Bad Wines:** The histogram has a higher peak than the normal distribution and is slightly right-skewed compared to the normal density.

## Exercise (b)

```{r}
pH.good <- working.df %>% filter(good == 1) %>% pull(pH)
pH.bad <- working.df %>% filter(good == 0) %>% pull(pH)
pH.all <- working.df %>% pull(pH)

#QQ-Plots:
ggplot(data.frame(pH = pH.good), aes(sample = pH)) +
  geom_qq() +
  geom_qq_line(col = "red") +
  labs(x = "Theoretical Quantiles",
       y = "Empirical Quantiles",
       title = "QQ: pH of Good Wines against Normal Distr.")

ggplot(data.frame(pH = pH.bad), aes(sample = pH)) +
  geom_qq() +
  geom_qq_line(col = "red") +
  labs(x = "Theoretical Quantiles",
       y = "Empirical Quantiles",
       title = "QQ: pH of Bad Wines against Normal Distr.")

ggplot(data.frame(pH = pH.all), aes(sample = pH)) +
  geom_qq() +
  geom_qq_line(col = "red") +
  labs(x = "Theoretical Quantiles",
       y = "Empirical Quantiles",
       title = "QQ: pH of All Wines against Normal Distr.")

#Data-Prep for PP-Plots
ecdf.pH.all <- ecdf(sort(pH.all))
norm.all <- pnorm(sort(pH.all), mean(pH.all), sd = sd(pH.all))

ecdf.pH.good <- ecdf(sort(pH.good))
norm.good <- pnorm(sort(pH.good), mean(pH.good), sd = sd(pH.good))

ecdf.pH.bad <- ecdf(sort(pH.bad))
norm.bad <- pnorm(sort(pH.bad), mean(pH.bad), sd = sd(pH.bad))

#PP-Plots
ggplot(data.frame(norm.all, ecdf.pH.all(sort(pH.all))), 
       aes(x = norm.all, y = ecdf.pH.all(sort(pH.all)))) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, color = "red") +
  labs(x = "Theoretical Probabilities",
       y = "Empirical Probabilities",
       title = "PP: pH of All Wines against Normal Distr.")

ggplot(data.frame(norm.good, ecdf.pH.good(sort(pH.good))), 
       aes(x = norm.good, y = ecdf.pH.good(sort(pH.good)))) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, color = "red") +
  labs(x = "Theoretical Probabilities",
       y = "Empirical Probabilities",
       title = "PP: pH of Good Wines against Normal Distr.")

ggplot(data.frame(norm.bad, ecdf.pH.bad(sort(pH.bad))), 
       aes(x = norm.bad, y = ecdf.pH.bad(sort(pH.bad)))) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, color = "red") +
  labs(x = "Theoretical Probabilities",
       y = "Empirical Probabilities",
       title = "PP: pH of Bad Wines against Normal Distr.")
```

QQ-plots compare the quantiles of the sample data to the quantiles of a theoretical distribution, in this case the normal distribution, whereas PP-plots compare the ecdfs of the sample data and the theoretical distribution. For both holds: If the data follows the theoretical distribution, the points should lie approximately along a straight line.

Here are some key observations that are also in line with what has already been mentioned in Exercise (a):

**Good Wines:**

-   **PP-Plot**: The PP-plot looks good, very close to the normal distribution, indicating a high goodness-of-fit overall.

-   **QQ-Plot**: There is an upwards curve in the tails, indicating heavier tails with more extreme values (outliers) than the theoretical distribution.

**Bad Wines:**

-   **PP-Plot**: The PP-plot shows some deviation from the straight line, indicating a less perfect fit.

-   **QQ-Plot**: Points deviate above the line in the upper tail but follow the line in the lower tail, suggesting that the sample data is right-skewed.

**All Wines:**

-   **PP-Plot**: Again, the PP-plot shows some deviation from the straight line, indicating a less perfect fit.

-   **QQ-Plot**: There are slightly heavy tails, more pronounced in the upper tail, likely due to the skewness of the bad wines.

## Exercise (c)

```{r}
# Function to calculate ecdf and confidence intervals
ecdf.with.ci <- function(data, alpha = 0.05) {
  n <- length(data)
  ecdf.data <- ecdf(data)
  x <- sort(unique(data))
  y <- ecdf.data(x)
  
  # Slutzky's lemma and general limit theorem: Using var of the sample to approx. real var
  epsilon <- sqrt(var(data) * log(2 / alpha))
  lower <- pmax(y - epsilon, 0)
  upper <- pmin(y + epsilon, 1)
  
  data.frame(x = x, y = y, lower = lower, upper = upper)
}

pH.good <- working.df %>% filter(good == 1) %>% pull(pH)
pH.bad <- working.df %>% filter(good == 0) %>% pull(pH)
pH.all <- working.df %>% pull(pH)

ecdf.all <- ecdf.with.ci(pH.all)
ecdf.good <- ecdf.with.ci(pH.good)
ecdf.bad <- ecdf.with.ci(pH.bad)

# Plot all wines
ggplot(ecdf.all, aes(x = x, y = y)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.3, fill = "blue") +
  geom_step(color = "red") +
  labs(title = "Empirical Distribution Functions of pH for All Wines",
       x = "pH", 
       y = "Cumulative Probability")

# Plot good wines
ggplot(ecdf.good, aes(x = x, y = y)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.3, fill = "blue") +
  geom_step(color = "red") +
  labs(title = "Empirical Distribution Functions of pH for Good Wines", 
       x = "pH", 
       y = "Cumulative Probability")

# Plot bad wines
ggplot(ecdf.bad, aes(x = x, y = y)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.3, fill = "blue") +
  geom_step(color = "red") +
  labs(title = "Empirical Distribution Functions of pH for Bad Wines", 
       x = "pH", 
       y = "Cumulative Probability")

```

## Exercise (d)

```{r}
# Function to calculate ecdf and uniform confidence intervals using KS method
ecdf.with.ks.ci <- function(data, alpha = 0.05) {
  n <- length(data)
  ecdf.data <- ecdf(data)
  x <- sort(unique(data))
  y <- ecdf.data(x)
  
  # Kolmogorov-Smirnov criterion for confidence bands
  epsilon <- sqrt(-0.5 * log(alpha / 2)) / sqrt(n)
  lower <- pmax(y - epsilon, 0)
  upper <- pmin(y + epsilon, 1)
  
  data.frame(x = x, y = y, lower = lower, upper = upper)
}

# From this line on: same code as in Exercise (c)

pH.good <- working.df %>% filter(good == 1) %>% pull(pH)
pH.bad <- working.df %>% filter(good == 0) %>% pull(pH)
pH.all <- working.df %>% pull(pH)

ecdf.all <- ecdf.with.ks.ci(pH.all)
ecdf.good <- ecdf.with.ks.ci(pH.good)
ecdf.bad <- ecdf.with.ks.ci(pH.bad)

# Plot all wines
ggplot(ecdf.all, aes(x = x, y = y)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.3, fill = "blue") +
  geom_step(color = "red") +
  labs(title = "Empirical Distribution Functions of pH for All Wines",
       x = "pH", 
       y = "Cumulative Probability")

# Plot good wines
ggplot(ecdf.good, aes(x = x, y = y)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.3, fill = "blue") +
  geom_step(color = "red") +
  labs(title = "Empirical Distribution Functions of pH for Good Wines", 
       x = "pH", 
       y = "Cumulative Probability")

# Plot bad wines
ggplot(ecdf.bad, aes(x = x, y = y)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.3, fill = "blue") +
  geom_step(color = "red") +
  labs(title = "Empirical Distribution Functions of pH for Bad Wines", 
       x = "pH", 
       y = "Cumulative Probability")
```

Remark: Calculation probably not right!

## Exercise (d)

```{r}
# Cell only works with the cells above run.
# Combine good and bad data for plotting
combined.data <- rbind(
  mutate(ecdf.good, type = "Good"),
  mutate(ecdf.bad, type = "Bad")
)

# Plot ECDF with confidence intervals for good and bad wines
ggplot(combined.data, aes(x = x, y = ecdf, color = type)) +
  geom_step() +
  geom_ribbon(aes(ymin = lower.ci, ymax = upper.ci), alpha = 0.2) +
  labs(title = "Empirical Distribution Functions of pH for Good and Bad Wines",
       x = "pH", 
       y = "Cumulative Probability",
       color = "Wine Type") +
  scale_color_manual(values = c("Good" = "blue", "Bad" = "red"))
```

Remark: calculation is not correct yet!

```{r}
# Function to calculate ECDF and pointwise confidence bands
ecdf_confidence_bands <- function(data, alpha = 0.05) {
  n <- length(data)
  ecdf_data <- ecdf(data)
  x <- sort(unique(data))
  y <- ecdf_data(x)
  
  epsilon <- sqrt((1/(2*n)) * log(2/alpha))
  lower <- pmax(y - epsilon, 0)
  upper <- pmin(y + epsilon, 1)
  
  data.frame(x = x, y = y, lower = lower, upper = upper)
}

# Calculate ECDF and confidence bands for each group
ecdf_good <- ecdf_confidence_bands(pH.good)
ecdf_bad <- ecdf_confidence_bands(pH.bad)
ecdf_all <- ecdf_confidence_bands(pH.all)

# Plotting
ggplot() +
  geom_step(data = ecdf_good, aes(x = x, y = y), color = 'blue', linetype = "solid", size = 1) +
  geom_ribbon(data = ecdf_good, aes(x = x, ymin = lower, ymax = upper), fill = 'blue', alpha = 0.2) +
  geom_step(data = ecdf_bad, aes(x = x, y = y), color = 'red', linetype = "solid", size = 1) +
  geom_ribbon(data = ecdf_bad, aes(x = x, ymin = lower, ymax = upper), fill = 'red', alpha = 0.2) +
  geom_step(data = ecdf_all, aes(x = x, y = y), color = 'green', linetype = "solid", size = 1) +
  geom_ribbon(data = ecdf_all, aes(x = x, ymin = lower, ymax = upper), fill = 'green', alpha = 0.2) +
  labs(title = "Empirical Distribution Functions of pH for Good, Bad, and All Wines",
       x = "pH", y = "ECDF",
       color = "Group") +
  scale_color_manual(values = c("blue" = "Good", "red" = "Bad", "green" = "All")) +
  theme_minimal()
```
